Asteroseismology of GD358 with complex C/O core profiles

### 2016-03-17 msnew.tex
Draft ready for review

\begin{abstract}
We report on the analysis of 33 years of photometric observations of the pulsating
helium atmosphere white dwarf GD358.  The complete data set includes archival data from 1982-2006, 
and 1105.1 hours of new observations from 2007-2015. From this data set, we extract 15 frequencies 
representing g-mode pulsation modes, adding 4 modes to the 11 modes 
known previously. We present evidence that these 15 modes are $\ell=1$ modes, 13 of which belong to 
a consecutive sequence in radial overtone
$k$. We perform a detailed asteroseismic analysis using models that include parameterized, complex carbon and oxygen 
core composition profiles to fit the periods. Recent spectroscopic analyses place GD358 near the red edge of the DBV instability strip, at 24,000 $\pm$ 500~K and a \logg of 7.8 $\pm$ 0.08 dex. The surface gravity translates to a mass range of 0.455 to 0.540 \msun. Our best fit model is slightly more massive and much hotter (by 2,000~K) than the most recent spectroscopy indicates. We find a pure helium layer mass of $10^{-5.40}$, consistent with a red edge object when compared with other DBVs analyzed using the same models.

@@

The basic method in our asteroseismic fitting consists of calculating grids of white dwarf models and 
running a fitting subroutine to match the periods of the models ($\rm P^{obs calc}$) with the observed 
periods $P^{\rm obs}$. Following standard statistical methods, each fit is assigned a fitness parameter calculated the following way:

@@

We started with a master grid (Table \ref{fitt1}) chosen so that it covered all relevant 
area of parameter space and had sufficient resolution to find any region of local minimum 
in the fitness parameter \citep{Bischoff-Kim11a,Bischoff-Kim14}. We used the minimum resolution 
that was computationally manageable. The master grid involved the computation 
of 10,483,200 models. We fit simultaneously all 15 periods, requiring all of them to be $\ell=1$ 
modes. A fitness map of this initial fit are shown in the left panel of Fig. \ref{ffit2}, and the 
parameters of the best fit model are listed in Table \ref{fitt1}.

@@

We show a contour map of the location of the models that best match the average period spacing of 
39.08 seconds in Fig. \ref{ffit2}, right panel. We place it side by side with a contour map showing 
the location of the best fit model in the same region of parameter space, based on the master grid 
fitting described in section \ref{grids}. Note how the best fit model falls right within the valley 
where the period spacings between GD358 and the models match. This should come as no surprise, as in 
order for 15 periods to fit reasonably well, the model period list should have a spacing similar to 
that of the observed period spectrum. The most recent spectroscopic determination of \citet{Koester2013} is far off the valley, at 24,000~K and $0.506^{+0.034)_{-0.051}}$ \msun. Previous spectroscopic determinations, such as that of \citet{Bergeron2011}, do fall within the swath where the period spacings match. We show the spectroscopic data point of \citet{Bergeron2011} in Fig. \ref{ffit2}. The best fit model from the period by period fit lies further down the valley.

%  \includegraphics[width=1.0\columnwidth]{ffit2.eps}
\begin{figure}
\epsscale{0.7}
\plotone{ffit2.eps}
\caption{
{\it Left panel}: Contour map showing the location of the best fit models. The quantity plotted in 
the mass-effective temperature plane is the fitness parameter defined in equation \ref{fiteq1}. 
{\it Right panel}: Contour map of the difference between the average 
periods spacing determined from GD 358's periods and the asymptotic period spacing of the models, 
in the mass-effective temperature plane. The scale is in 10s of seconds, so the worse matches 
pictured on the plot have $\left|\Delta P_{\rm obs}- \Delta P_{\rm calc}\right| \sim 5 \; \rm s$. We plotted the spectroscopic mass and temperature determinations of \citet{Bergeron2011} (higher mass), and that of \citet{Koester2013} (lower mass). The point without error bars corresponds to the best initial fit model. \label{ffit2}
}
\end{figure}

One can fit simultaneously the average period spacing and the individual periods 
formally while performing the fits by using some prescription to calculate the goodness of fit. 
This leads to a more complex relation than defined in equation \ref{fiteq1}. 
Note that the period spacing is a much weaker constraint than the individual period fit. 
If one takes 5.0 seconds as an upper limit for goodness of fit, that includes 4\% of the points in the 
period-by-period fit plot (left panel in Fig \ref{ffit2}), but the entire parameter plane 
for the average period spacing fit plot (right panel). In our refined fits, we limit ourselves to a region very near the best fit model found on the initial grid. This essentially limits our search to models that already match the average period spacing. We avoid sophisticated schemes to calculate the fitness parameter and simply use equation \ref{fiteq1}.

@@

One way to gain a sense of how fine the grid needs to be is to make single 
parameter cuts through parameter space. Fig.  \ref{ffit3} shows such cuts 
for master grid models. The plot was made by fixing 5 of the 6 parameters 
to the best fit values of the best initial fit (see table \ref{fitt1}).
For some parameters, the fits seem to settle to a minimum in a 
smooth way, while for others, they exhibit jumps. For instance, the spike in the effective temperature plot at 27,800~K is due to a period (around 530~s) that goes away and then comes back. The model with the missing period fits poorly compared to the models on either side of it. The jump from \sigrms $\sim$ 5 to \sigrms $\sim$ 30 that happens from 28,400~K to 28,600~K is due to a discontinuous change in the period spectrum of the models. Namely, the appearance of a new mode between the 710 and 784~s modes of the 28,400~K model.

\begin{figure}
\epsscale{1.0}
\plotone{ffit3.eps}
%  \includegraphics[scale=0.70,viewport=50 0 50 350]{f2.png}
%  \includegraphics[width=0.9\columnwidth]{ffit3.eps}
\caption{
Dependence of the fitness parameter on different parameters. The 5 parameters other than the one shown 
on the horizontal axis in each plot are held to fixed values corresponding to the initial best fit model (table \ref{ffit1}). The vertical dashed lines mark the ranges considered in the refined fitting. For the envelope mass, we went up to \menv = 2.3.
 \label{ffit3}
}
\end{figure}

Discontinuities like these are unsettling, as it is easy to see that one might miss a best fit 
model if the grid is not fine enough. In order to quantify "fine enough", 
we ran systematic scans, computing models with 5 parameter fixed and allowing 
the 6th parameter to vary in very fine steps. We went down in 
step sizes to $\Delta {\rm T_{eff}}=1$~K, $\Delta M_*=0.0001 M_\odot$, 
$\Delta M_{\rm env}= \Delta M_{\rm He}=0.01$ dex, $\Delta X_o=0.01$, and $\Delta q_{\rm fm}=0.001$. 
Regardless of the behavior of individual modes in the models as the parameters vary, 
we can assert what step sizes will allow us to minimize our risk 
of missing a minimum. We settled on step sizes of 
$\Delta {\rm T_{eff}}=50$, $\Delta M_*=0.0001$, $\Delta M_{\rm env}=0.1$, $\Delta M_{\rm He}=0.1$, 
$\Delta X_o=0.1$, and $\Delta q_{\rm fm}=0.005$ for our refined grid.

@@

The parameters for our best fit model based on the refined grid are listed in Table \ref{fitt1} and the periods of that model in 
Table \ref{gd}. The goodness of fit of the model is \sigrms$=0.964$~s. We also list
the Bayes Information Criterion (BIC) number, a statistic that
normalizes the quality of fits by number of free parameters and
number of constraints for comparison with other studies. For a discussion applied to this parameter
study, see \citet{Bischoff-Kim11a}. A negative BIC indicates a good quality of fit, given the number of constraints. One should keepp in mind that the quality of fit is aided by the fact that some of the periods have large uncertainties.

@@

We first performed a fitting of the periods using the master (coarse) grid described in section \ref{grids}. This 
placed the best fit model in the appropriate region of parameter space. Then we refined our fits, using the grids 
described at the end of section \ref{periodspacing}. We are able to recover the best fit parameters adequately, with 
the all top 5 best fit models having parameters 
${\rm T_{eff}}=25000$~K, \mhe $=-2.1$, $M_{\rm env}=-5.5$, $X_o=0.50$, and $q_{\rm fm}=0.190$ and a mass ranging between  0.5730 and 0.5734 \msun. The best fit model has \sigrms $=0.31$~s. We remind the reader that 
the step sizes in the second phase of the fitting are 50~K for ${\rm T_{eff}}$, 0.0001 \msun for stellar mass, 0.1 for 0.1 for helium layer masses, 0.1 for $X_o$, and 0.005 for $q_{\rm fm}$.

@@

We analyzed archival data and over a thousand hours of new observations on GD358, together covering a span of 33 years. This allowed us to extract 15 frequencies. This adds 4 modes to the 11 modes known previously. We performed a new asteroseismic fit of GD358 with the new modes and with models that include carbon and oxygen core composition profiles based on the stellar evolution models of \citet{Salaris97}.

WE find a best fit effective temperature of 23,650~K and a mass of 0.5706 \msun for GD358. While the temperature is close to the recent spectroscopic determination of 24000~K \citep{Koester2013}, the mass is over one sigma above the spectroscopic mass. On the other hand, the mass matches almost exactly that found by \citet{Bergeron2011} (but the effective temperature is low). The fact that the spectroscopic data point of \citet{Koester2013} is far off the asymptotic period spacing trend (Fig. \ref{ffit2}, right panel) is a concern.

In fitting GD358, we discovered that the best fit parameter space was semi-chaotic, with two very close sets of parameters able to yield very different fitness number. This is due to the fact that some periods can change values drastically for a small change in a given parameter. The phenomenon is only semi-chaotic, as the period will oscillate between two values. This is good news, because there remains a well defined minimum in parameter space, even if it is a double valued minimum. Any search algorithm that uses a small enough mesh should be able to find the minimum. This phenomenon likely became apparent while fitting GD358 because it is the first white dwarf for which we fit a long consecutive sequence of modes of the same $\ell$ with high resolution grids.

In this study, we varied the parameters that have traditionally been varied in this type of asteroseismic fitting, so that we can place our results side by side with other studies of DBVs, including KIC 8626021 \citep{Bischoff-Kim14}, EC200058 \citep{Bischoff-Kim11a}, and CBS114 \citep{Metcalfe05a}. We now have 4 DBVs that were the object of asteroseismic fitting that used a consistent set of models. Among the 4, cooler best fit models do exhibit thicker pure helium envelopes. Pinning down the proper effective temperature of each star is challenging. To be fully consistent, it is best to use the asteroseismic values, even if effective temperature is not strongly constrained. We show the relationship between asteroseismic effective temperature and pure helium layer mass in Fig. \ref{ffit5}. Stellar mass also comes into play and ideally, we would like all 4 models to have the same mass. The models range in mass from 0.525 (Ec20058) to 0.640 \msun (CBS114). More detailed stellar evolution calculations would have to be performed in order to assess the significance of the trend observed with these 4 DBVs.

@@

Numerical experiments of the type presented in section \ref{refinedfits} have shown that, consistent with 
the theory of non-radial oscillations, the shapes of the transition zones matter as much as where the transition zones occur \citep{Bischoff-Kim15}. Modern asteroseismic fitting vary parameters attached to the shape of transition zones \citep{Giammichele15}. We refrained from doing this in the present study because we wanted to compare our results with previous studies of DBVs. It is unclear how much of an effect on structure parameters a change in parameterization would have.

With data spanning a long period of time, we learn about the stability of the different modes. We find that the shorter period modes tend to be more stable, while the longer period modes tend to vary more in frequency over time. \Citet{Bell15} have observed and modeled such stochastic behavior in KIC 4552982, a red edge DAV observed nearly continuously for 1.5 years by \emph{Kepler}. In that star, the 361.58 s triplet has sharply defined peaks, while the rest of the modes, with much higher overtone numbers, are less stable. The stability of the low k modes is likely due to the fact that they are strongly trapped in the core. Unlike their higer k counterparts, they are affected very little by the convection zone near the surface.

>>>>

\begin{abstract}
We report on the analysis of 33 years of photometric observations of the pulsating
helium atmosphere white dwarf GD358.  The complete data set includes archival data from 1982-2006, 
and 1105.1 hours of new observations from 2007-2015. From this data set, we extract 15 frequencies 
representing g-mode pulsation modes, adding 4 modes to the 11 modes 
known previously. We present evidence that these 15 modes are $\ell=1$ modes, 13 of which belong to 
a consecutive sequence in radial overtone
$k$. We perform a detailed asteroseismic analysis using models that include parameterized, complex carbon and oxygen 
core composition profiles to fit the periods. Recent spectroscopic analyses place GD358 near the red edge of the DBV instability strip, at 24,000 $\pm$ 500~K and a \logg of 7.8 $\pm$ 0.08 dex. The surface gravity translates to a mass range of 0.455 to 0.540 \msun. Our best fit model has a temperature of 23,650~K and a mass of 0.5706 \msun ~. That is slightly more massive than what the most recent spectroscopy suggests. We find a pure helium layer mass of $10^{-5.50}$, consistent with the outward diffusion of helium over time.

@@

The basic method in our asteroseismic fitting consists of calculating grids of white dwarf models and 
running a fitting subroutine to match the periods of the models ($\rm P^{calc}$) with the observed 
periods $P^{\rm obs}$. Following standard statistical methods, each fit is assigned a fitness parameter calculated the following way:

@@

We started with a master grid (Table \ref{fitt1}) chosen so that it covered all relevant 
area of parameter space and had sufficient resolution to find any region of local minimum 
in the fitness parameter \citep{Bischoff-Kim11a,Bischoff-Kim14}. We used the maximum resolution 
that was computationally manageable. The master grid involved the computation 
of 10,483,200 models. We fit simultaneously all 15 periods, requiring all of them to be $\ell=1$ 
modes. A fitness map of this initial fit are shown in the left panel of Fig. \ref{ffit2}, and the 
parameters of the best fit model are listed in Table \ref{fitt1}.

@@

We show a contour map of the location of the models that best match the average period spacing of 
39.08 seconds in Fig. \ref{ffit2}, right panel. We place it side by side with a contour map showing 
the location of the best fit model in the same region of parameter space, based on the master grid 
fitting described in section \ref{grids}. Note how the best fit model falls right within the valley 
where the period spacings between GD358 and the models match. This should come as no surprise, as in 
order for 15 periods to fit reasonably well, the model period list should have a spacing similar to 
that of the observed period spectrum. The most recent spectroscopic determination of \citet{Koester2013} is far off the valley, at 24,000~K and $0.506^{+0.034}_{-0.051}$ \msun. Previous spectroscopic determinations, such as that of \citet{Bergeron2011}, do fall within the swath where the period spacings match.

%  \includegraphics[width=1.0\columnwidth]{ffit2.eps}
\begin{figure}
\epsscale{1.0}
\plotone{ffit2.eps}
\caption{
{\it Left panel}: Contour map showing the location of the best fit models. The quantity plotted in 
the mass-effective temperature plane is the fitness parameter defined in equation \ref{fiteq1}. 
{\it Right panel}: Contour map of the difference between the average 
periods spacing determined from GD 358's periods and the asymptotic period spacing of the models, 
in the mass-effective temperature plane. The scale is in 10ths of seconds, so the worse matches 
pictured on the plot have $\left|\Delta P_{\rm obs}- \Delta P_{\rm calc}\right| \sim 5 \; \rm s$. We plotted the spectroscopic mass and temperature determinations of \citet{Bergeron2011} (higher mass), and that of \citet{Koester2013} (lower mass). The point without error bars corresponds to the best initial fit model. \label{ffit2}
}
\end{figure}

One can fit simultaneously the average period spacing and the individual periods 
formally while performing the fits by using some prescription to calculate the goodness of fit. 
This leads to a more complex relation than defined in equation \ref{fiteq1}. 
Note that the period spacing is a much weaker constraint than the individual period fit. 
If one takes 5.0 seconds as an upper limit for goodness of fit, that includes 4\% of the models in the 
period-by-period fit plot (left panel in Fig \ref{ffit2}), but the entire parameter plane 
for the average period spacing fit plot (right panel). In our refined fits, we limit ourselves to a region very near the best fit model found on the initial grid. This essentially limits our search to models that already match the average period spacing. We avoid sophisticated schemes to calculate the fitness parameter and simply use equation \ref{fiteq1}.

@@

One way to gain a sense of how fine the grid needs to be is to make single 
parameter cuts through parameter space. Fig.  \ref{ffit3} shows such cuts 
for master grid models. The plot was made by fixing 5 of the 6 parameters 
to the best fit values of the best initial fit (see table \ref{fitt1}).
For some parameters, the fits seem to settle to a minimum in a 
smooth way, while for others, they exhibit jumps. For instance, the spike in the effective temperature plot at 27,800~K is due to a period (around 530~s) that goes away and then comes back. The model with the missing period fits poorly compared to the models on either side of it. The jump from \sigrms $\sim$ 5 s to \sigrms $\sim$ 30 s that happens from 28,400~K to 28,600~K is due to a discontinuous change in the period spectrum of the models. Namely, the appearance of a new mode between the 710 and 784~s modes of the 28,400~K model.

\begin{figure}
\epsscale{1.0}
\plotone{ffit3.eps}
%  \includegraphics[scale=0.70,viewport=50 0 50 350]{f2.png}
%  \includegraphics[width=0.9\columnwidth]{ffit3.eps}
\caption{
Dependence of the fitness parameter on different parameters. The 5 parameters other than the one shown 
on the horizontal axis in each plot are held to fixed values corresponding to the initial best fit model (table \ref{ffit1}). The vertical dashed lines mark the ranges considered in the refined fitting. For the envelope mass, we went from \menv = -2.0 to -2.3.
 \label{ffit3}
}
\end{figure}

Discontinuities like these are unsettling, as it is easy to see that one might miss a best fit 
model if the grid is not fine enough. In order to quantify "fine enough", 
we ran systematic scans, computing models with 5 parameter fixed and allowing 
the 6th parameter to vary in very fine steps. We went down in 
step sizes to $\Delta {\rm T_{eff}}=1$~K, $\Delta M_*=0.0001 M_\odot$, 
$\Delta M_{\rm env}= \Delta M_{\rm He}=0.01$ dex, $\Delta X_o=0.01$, and $\Delta q_{\rm fm}=0.001$. 
Regardless of the behavior of individual modes in the models as the parameters vary, 
we can assert what step sizes will allow us to minimize our risk 
of missing a minimum, while minimizing the number of models to compute. We settled on step sizes of 
$\Delta {\rm T_{eff}}=50$~K, $\Delta M_*=0.0001 M_\odot$, $\Delta M_{\rm env}=\Delta M_{\rm He}=0.1$ dex,  
$\Delta X_o=0.1$, and $\Delta q_{\rm fm}=0.005$ for our refined grid.

@@

The parameters for our best fit model based on the refined grid are listed in Table \ref{fitt1} and the periods of that model in 
Table \ref{gd}. The goodness of fit of the model is \sigrms$=0.964$~s. We also list
the Bayes Information Criterion (BIC) number, a statistic that
normalizes the quality of fits by number of free parameters and
number of constraints for comparison with other studies. For a discussion applied to this parameter
study, see \citet{Bischoff-Kim11a}. A negative BIC indicates a good quality of fit, given the number of constraints. One should keepp in mind that the quality of fit is aided by the fact that some of the observed periods have large uncertainties.

@@

We first performed a fitting of the periods using the master (coarse) grid described in section \ref{grids}. This 
placed the best fit model in the appropriate region of parameter space. Then we refined our fits, using the grids 
described at the end of section \ref{periodspacing}. We are able to recover the best fit parameters adequately, with 
the all top 5 best fit models having parameters 
${\rm T_{eff}}=25000$~K, \mhe $=-2.1$, $M_{\rm env}=-5.5$, $X_o=0.50$, and $q_{\rm fm}=0.190$ and a mass ranging between  0.5730 and 0.5734 \msun. The best fit model has \sigrms $=0.31$~s. We remind the reader that 
the step sizes in the second phase of the fitting are 50~K for ${\rm T_{eff}}$, 0.0001 \msun ~ for stellar mass, 0.1 dex for helium layer masses, 0.1 for $X_o$, and 0.005 for $q_{\rm fm}$.

@@

We analyzed archival data and over a thousand hours of new observations on GD358, together covering a span of 33 years. With data spanning such a long period of time, we learn about the stability of the different modes. We find that the shorter period modes tend to be more stable, while the longer period modes tend to vary more in frequency over time. \citet{Bell15} have observed and modeled such stochastic behavior in KIC 4552982, a red edge DAV observed nearly continuously for 1.5 years by \emph{Kepler}. In that star, the 361.58 s triplet has sharply defined peaks, while the rest of the modes, with much higher overtone numbers, are less stable. The stability of the low k modes is likely due to the fact that they are strongly trapped in the core. Unlike their higer k counterparts, they are affected very little by the convection zone near the surface.

In analyzing the data, we found 4 new modes, adding to the 11 modes known previously. With these 15 modes, we performed a new asteroseismic fit of GD358 with models that include carbon and oxygen core composition profiles based on the stellar evolution models of \citet{Salaris97}. We find a best fit effective temperature of 23,650~K and a mass of 0.5706 \msun ~ for GD358. While the temperature is close to the recent spectroscopic determination of 24000~K \citep{Koester2013}, the mass is over one sigma above the spectroscopic mass. On the other hand, the mass matches almost exactly that found by \citet{Bergeron2011} (but the effective temperature is low). The spectroscopic data point of \citet{Koester2013} is somewhat off the asymptotic period spacing trend (Fig. \ref{ffit2}, right panel).

In this study, we varied the parameters that have traditionally been varied in this type of asteroseismic fitting, so that we can place our results side by side with other studies of DBVs, including KIC 8626021 \citep{Bischoff-Kim14}, EC200058 \citep{Bischoff-Kim11a}, and CBS114 \citep{Metcalfe05a}. We now have 4 DBVs that were the object of asteroseismic fitting that used a consistent set of models. We find that cooler best fit models have thicker pure helium envelopes (Fig. \ref{ffit5}), in accordance with the outward diffusion of helium over time. To be consistent, we used the effective temperature found through the asteroseismic fitting of each star. Stellar mass also comes into play in diffusion. Ideally, we would like all 4 models to have the same mass. The models range in mass from 0.525 (EC20058) to 0.640 \msun ~ (CBS114). More detailed stellar evolution calculations would have to be performed in order to assess the significance of the trend observed with these 4 DBVs.

@@

Numerical experiments of the type presented in section \ref{refinedfits} have shown that, consistent with 
the theory of non-radial oscillations, the shapes of the transition zones matter as much as where they occur \citep{Bischoff-Kim15}. Modern asteroseismic fitting vary parameters attached to the shape of transition zones \citep{Giammichele15}. We refrained from doing this in the present study because we wanted to compare our results with previous studies of DBVs. It is unclear how much of an effect on structure parameters a change in parameterization would have.

In fitting GD358, we discovered that the best fit parameter space was semi-chaotic, with two very close sets of parameters able to yield very different fitness parameters. This is due to the fact that some periods can change values drastically for a small change in a given parameter. The phenomenon is only semi-chaotic, as the period will oscillate between two values. This is good news, because there remains a well defined minimum in parameter space, even if it is a double valued minimum. Any search algorithm that uses a small enough mesh should be able to find the minimum. This phenomenon likely became apparent while fitting GD358 because it is the first white dwarf for which we fit a long consecutive sequence of modes of the same $\ell$ with high resolution grids.
